---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
  fig.path = "man/figures/README-", out.width = "100%"
)
```

# remotePARTS

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![R-CMD-check](https://github.com/morrowcj/remotePARTS/workflows/R-CMD-check/badge.svg)](https://github.com/morrowcj/remotePARTS/actions)
[![status](https://joss.theoj.org/papers/c6a3da6a56aa0fb0e1f8a4f36cab12c2/status.svg)](https://joss.theoj.org/papers/c6a3da6a56aa0fb0e1f8a4f36cab12c2)

<!-- badges: end -->

`remotePARTS` is an software package for the `R` statistical programming language. 
The package contains tools for analyzing spatiotemporal data, typically obtained 
via remote sensing.

## Description

These tools were created to test map-scale hypotheses about trends in large 
remotely sensed data sets, but they are useful for analyzing trends in any 
spatial data, with or without a temporal compoenent. Statistical tests are 
conducted with the PARTS method for analyzing spatially autocorrelated time 
series (Ives et al., 2021). The method's unique approach can handle extremely
large data sets that other spatiotemporal models cannot, while still 
appropriately accounting for spatial and temporal autocorrelation. This is 
done by partitioning the data into smaller chunks, analyzing chunks separately
and then combining the separate analyses into a single test, that acconts for
correlations among chunks, of the map-scale hypotheses.

## Instalation

To install the package and it's dependencies from CRAN, use the following R code:

```{r, eval = FALSE}
install.packages("remotePARTS")
```

To install the latest stable version of this package from github, use

```{r, eval = FALSE}
remotes::install_github("morrowcj/remotePARTS")
```

and to test out the newest features and functionality, use 

```{r, eval = FALSE}
remotes::install_github("morrowcj/remotePARTS", ref = "develop")
```
Then, upon successful installation, load the package with `library(remotePARTS)`. 

### Dependencies

Since the matrix operations in this package rely on `C++` code, as implemented 
via the [RcppEigen package](https://github.com/RcppCore/RcppEigen),
the latest version of [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
is required for Windows and C++11 is required for other systems. 

<!--- Now that the package is on CRAN, I am not 100% certain that the above statement
is strictly true anymore, but I'm going to leave it for now until I learn more. --->

## Example usage

<!-- TODO: This whole section needs to be reworked --->

For examples on how to use `remotePARTS`, see the `Alaska` vignette:

```{r, eval = FALSE}
vignette("Alaska")
```

Note that the vignette needs to be built when installing with and may require 
the `build_vignettes = TRUE` argument when installing with `install_github()`.

If you're having trouble installing or building the package, you may need to 
double check that the R build tools are properly installed on your
machine: 
[official Rstudio development prerequisites](https://support.posit.co/hc/en-us/articles/200486498-Package-Development-Prerequisites)
To do this, use `pkgbuild::has_build_tools(debug = TRUE)` and 
`pkgbuild::check_build_tools(debug = TRUE)` to unsure that your build tools are
up to date.

The lastest stable version of the vignette is also hosted online at 
https://morrowcj.github.io/remotePARTS/Alaska.html.

## Contribution, bugs, and feature requests

If you wish to contribute to this package, report bugs, suggest new features
or behavior, correct typos, update documentation, or anything else, please submit a
[GitHub Issue](https://github.com/morrowcj/remotePARTS/issues). We welcome and
appreciate any and all feedback.

# References

Ives, Anthony R., et al. "Statistical inference for trends in spatiotemporal data."
Remote Sensing of Environment 266 (2021): 112678. https://doi.org/10.1016/j.rse.2021.112678 
