---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
  fig.path = "man/figures/README-", out.width = "100%"
)
```

# remotePARTS

<!-- badges: start -->
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

<!-- [![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing) -->

[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)

<!-- [![Travis build status](https://travis-ci.com/morrowcj/remotePARTS.svg?branch=master)](https://travis-ci.com/morrowcj/remotePARTS) -->

<!-- [![Travis build status](https://travis-ci.com/morrowcj/remotePARTS.svg?branch=master)](https://travis-ci.org/github/morrowcj/remotePARTS) -->


[![R-CMD-check](https://github.com/morrowcj/remotePARTS/workflows/R-CMD-check/badge.svg)](https://github.com/morrowcj/remotePARTS/actions)
<!-- badges: end -->

`remotePARTS` is an `R` package that contains tools for analyzing spatiotemporal
data, typically obtained via remote sensing. 

## Description

These tools were created to test map-scale hypotheses about trends in large 
remotely sensed data sets but any data with spatial and temporal variation
can be analyzed. Tests are conducted using the PARTS method for analyzing spatially
autocorrelated time series (Ives et al., 2021). The method's unique approach can
handle extremely large data sets that other spatiotemporal models cannot, while
still appropriately accounting for spatial and temporal autocorrelation. This 
is done by partitioning the data into smaller chunks, analyzing chunks separately
and then combining the separate analyses into a single, correlated test of the 
map-scale hypotheses.

## Instalation

To install the package and it's dependencies, use the following R code:

```{r, eval = FALSE}
install.packages("remotePARTS")
```

To install the latest development version of this package from github, use

```{r, eval = FALSE}
install.packages("devtools") # ensure you have the latest devtools
devtools::install_github("morrowcj/remotePARTS")
```

Then, upon successful installation, load the package with `library(remotePARTS)`. 

The latest version of [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
is required for Windows and C++11 is required for other systems. 

## Example usage

For examples on how to use `remotePARTS`, see the `Alaska` vignette:

```{r, eval = FALSE}
vignette("Alaska")
```

Note that the vignette needs to be built when installing with and may require 
the `build_vignettes = TRUE` argument when installing with `install_github()`.

If you're having trouble installing or building the package, you may need to 
double check that the R build tools are properly installed on your
machine: [official Rstudio development prerequisites](https://support.rstudio.com/hc/en-us/articles/200486498-Package-Development-Prerequisites)
To do this, use `pkgbuild::has_build_tools(debug = TRUE)` and 
`pkgbuild::check_build_tools(debug = TRUE)` to unsure that your build tools are
up to date.

The vignette is also available online: 
https://morrowcj.github.io/remotePARTS/Alaska.html.

## Bugs and feature requests

If you find any bugs, have a feature or improvement to suggest, or any other feedback about the `remotePARTS` package,
please submit a GitHub Issue [here](https://github.com/morrowcj/remotePARTS/issues). We really appreciate any and all
feedback.

# References

Ives, Anthony R., et al. "Statistical inference for trends in spatiotemporal data." Remote Sensing of Environment 266 (2021): 112678. https://doi.org/10.1016/j.rse.2021.112678 
