---
title: "Compare remoteSTAR with Tony's code"
author: "Clay Morrow"
date: "8/30/2020"
output: html_document
---

This document was knitted on `r Sys.time()` and compares the original code
to the code found in the `remoteSTAR` package. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, 
                      warning = FALSE, collapse = TRUE, comment = "#>")
```

```{r}
source("remote_sensing_tools_8Jul20.R")

library(remoteSTAR)
```

```{r}
load("../data-backup/ndvi_AK3000.RData", verbose = TRUE)
df <- ndvi_AK3000
rm(ndvi_AK3000)
```
### Data setup

```{r}
# setup

library(geosphere)

df$land <- droplevels(df$land) 
X <- as.matrix(df[, -c(1:6)])
n = nrow(X); n.obs = ncol(X) 
location = df[, c("lng", "lat")] 
time.int = 1:n.obs 
time.scaled = (time.int-min(time.int))/max(time.int)
Dist = distm(location)/1000
fun = "exponential-power"
```

### CLS

```{r CLS}
CLS <- CLS.fit(X, time.scaled)
CLS.star <- cls_star(X, time.scaled)

## Are these equivalent?
all(all.equal(CLS$c, CLS.star$Est),
    all.equal(CLS$t, CLS.star$t),
    all.equal(CLS$b, CLS.star$x_t0.EST)
    )

Y = with(CLS.star, Est/mean)
```
### SpatialCor

```{r fitSpatCor one}
r.est <- spatialcor.fit.data(X = X, t.scale = time.scaled, data=df, 
                             fit.n.sample=500,
                             plot.fig=FALSE, FUN=fun)

r.est.star <- fit_spatialcor(X = X, time.scaled, fit.n = 500, fun = fun, 
                             location = location, scale.dist = TRUE)

rbind("tony" = r.est$spatialcor, "star" = r.est.star$spatialcor) ## close - differences due to sampling
```
### Vfit

```{r Vfit}
r = r.est.star$spatialcor

V <- V.fit(Dist, r, fun)
V.star <- fitV(Dist, r, fun)

all.equal(V, V.star)
```

### fit nugget

```{r nugget}
nugget <- nugget.fit(formula = 'Y ~ 0 + land', df, V, 
                     nugget.tol = 0.0001, verbose = FALSE)

mod.mat <- model.matrix(formula("Y ~ 0 + land"), data = df)
nugget.star <- optimizeNugget_cpp(X = mod.mat, V = V, y = Y, lower = 0, 
                                  upper = 1, tol = 0.0001)

all.equal(nugget, nugget.star)
```
### invChol

```{r invchol}
Vn <- (1 - nugget) * V + nugget * diag(nrow(V))
invcholVn <- t(backsolve(chol(Vn), diag(nrow(Vn))))

invchol.star <- tinvchol_cpp(V, nugget)

all.equal(invcholVn, invchol.star)
```
### GLS

```{r GLS}
gls <- GLS.fit(formula = "Y ~ 0 + land", data = df, invcholV=invchol.star,
               V = Vn)

X0 <- matrix(1, nrow = n)
gls.star <- fitGLS(mod.mat, V, Y, X0, nugget = nugget)

all(all.equal(unname(gls$coef), gls.star$betahat),
    all.equal(gls$t, gls.star$tstat),
    all.equal(gls$F, gls.star$Fstat),
    all.equal(c(gls$df1.F, gls$df2.F), gls.star$df.F)
    )

cbind(coef = gls$coef, pval = gls$p.t)

gls$p.F
```

It looks like none of the t-tests for land class were significant 
but the F-test is significant (with both methods)
when using land cover as the predictor as the predictor. 

### GLS partitioned

```{r glspart}
gls.part <- fitGLS.partition_rcpp(mod.mat, Y, X0, Dist, r, fun, 
                                  npart = 5, mincross = 4)
remoteSTAR::GLS.partition.pvalue(gls.part, nboot = 2000)
```

and the correlated F-test corroborates these results.

### Again without outliers

[to check if outlier removal caused discrepancy]

```{r remove outlr, eval = FALSE}
# remove potential outliers
outl <- which(abs(scale(Y)) > -qnorm(p = 1/n/10))
df.o <- df[-outl, ]
Y.o <- Y[-outl]
X.o <- X[-outl, ]
location.o <- location[-outl, ]
Dist.o <- Dist[-outl, -outl]

## and do everything again
r.est.o <- spatialcor.fit.data(X = X.o, t.scale = time.scaled, data=df.o, 
                               fit.n.sample=500, plot.fig=FALSE, FUN=fun)

r.est.o.star <- fit_spatialcor(X = X.o, time.scaled, fit.n = 500, fun = fun, 
                               location = location.o, scale.dist = TRUE)

rbind("tony" = r.est.o$spatialcor, "star" = r.est.o.star$spatialcor)
#> very close

r.o = r.est.o.star$spatialcor

V.o <- V.fit(Dist.o, r.o, fun)
V.o.star <- fitV(Dist.o, r.o, fun)

all.equal(V.o, V.o.star) 
#> TRUE

nugget.o <- nugget.fit(formula = 'Y.o ~ 0 + land', data = df.o, V = V.o,
                     nugget.tol = 0.0001, verbose = FALSE) 

mod.mat.o <- model.matrix(formula("Y.o ~ 0 + land"), data = df.o)
nugget.o.star <- optimizeNugget_cpp(X = mod.mat.o, V = V.o, y = Y.o, lower = 0, 
                                  upper = 1, tol = 0.0001)

all.equal(nugget.o, nugget.o.star) 
#> TRUE

Vn.o <- (1 - nugget.o) * V.o + nugget.o * diag(nrow(V.o))
invcholVn.o <- t(backsolve(chol(Vn.o), diag(nrow(Vn.o))))

invchol.o.star <- tinvchol_cpp(V.o, nugget.o)

all.equal(invcholVn.o, invchol.o.star) 
#> TRUE

gls.o <- GLS.fit(formula = "Y.o ~ 0 + land", data = df.o, 
                 invcholV=invchol.o.star, V = Vn.o)

X0.o <- matrix(1, nrow = nrow(X.o))
gls.o.star <- fitGLS(mod.mat.o, V.o, Y.o, X0.o, nugget = nugget.o)

all(all.equal(unname(gls.o$coef), gls.o.star$betahat),
    all.equal(gls.o$t, gls.o.star$tstat),
    all.equal(gls.o$F, gls.o.star$Fstat),
    all.equal(c(gls.o$df1.F, gls.o$df2.F), gls.o.star$df.F)
    ) 
#> TRUE

gls.o$p.F 
#> 0.003956
```

