---
title: "Alaska"
subtitle: "NDVI analysis {remoteSTAR}"
author: "Clay Morrow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alaska}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 5, fig.asp = 1, cache = FALSE,
  cache.extra = rand_seed # attempt to cache random seed for reproduction 
)
```

```{r setup, echo = FALSE}
options(digits = 4, scipen = 1)
library(remoteSTAR) # load the package
```

```{r random seed}
set.seed(58) # set the random seed for this document
```

```{r data setup, echo = FALSE, eval = FALSE}
## Full data ----
# read in the data
ndvi <- data.table::fread(file = "../data-raw/north_america_checked.csv")

# convert $land column to a factor
land.classes = c("Evergr needle","Evergr broad","Decid needle","Decid broad",
                 "Mixed forest","Shrubland","Savanna","Grassland","Cropland",
                 "Cropland mosaics") # all land classes (in order of lvs)

ndvi$land <- factor(ndvi$land, labels = land.classes)

# save the rdata file
if(!file.exists("../data/ndvi.rda")){
  save(ndvi, file =  "../data/ndvi.rda", compress = "xz")
}

## Alaska subset ----
ndvi_AK <- ndvi[ndvi$lng < -141, ] #North Am. west of -141 is approx. AK
ndvi_AK$land <- droplevels(ndvi_AK$land)
 
# relative contirbution of each land class to AK
tmp <- aggregate(
  x = ndvi_AK$land,
  by = list(landclass = ndvi_AK$land),
  FUN = function(x)length(x)/nrow(ndvi_AK)
)

# any land class that occurs less than 2% is rare
ndvi_AK$rare.land <- tmp[match(ndvi_AK$land, tmp$landclass), "x"] <= 0.02

# reorder columns so ndvi are all at the end
ord <- c(grep("ndvi", names(ndvi_AK), invert = TRUE), #columns without 'ndvi'
         grep("ndvi", names(ndvi_AK))) # cols with ndvi
ndvi_AK <- ndvi_AK[, ..ord] #..x is data.table notation

# save AK data file
if(!file.exists("../data/ndvi_AK.rda")){
  save(ndvi_AK, file = "../data/ndvi_AK.rda", compress = "xz")
}

## subset AK further
ndvi_AK3000 <- ndvi_AK[!ndvi_AK$rare.land, ] # only common land classes
ndvi_AK3000 <- ndvi_AK3000[sample.int(n = nrow(ndvi_AK3000), size = 3000), ] # 3000 pts

# save small AK data file
if(!file.exists("../data/ndvi_AK3000.rda")){
  save(ndvi_AK3000, file = "../data/ndvi_AK3000.rda", compress = "xz")
}
```

## Alaska 3000

The `ndvi_AK3000` data set contains NDVI remote sensing data for Alaska from a 
random 3000 pixels over the time period of 1982 through 2013 as a `data.table`.

**More info TBD**

First, we'll extract only the NDVI columns (`X`) and scale and center the time 
points:


```{r ndvi_AK3000 data}
# load the dataset
data("ndvi_AK3000")

# extract some useful info
X <- as.matrix(ndvi_AK3000[, -c(1:6)]) # columns containing NDVI measurements
n = nrow(X); p = ncol(X) # dimensions of X
location = ndvi_AK3000[, c("lng", "lat")] # geographic location of each site
time.int = 1:p # time points as standard integers
time.scaled = scale(time.int) # scaled and centered time
```

### Analysis of NDVI at the pixel level

Next we will get the average change in NDVI over time at *each site* (`fm.cls$Est`) 
and then scale those regression coefficients `rel.est`:

```{r cls_star AK3000}
# perform time series analysis for each site (no intercept)
fm.cls <- cls_star(X, t = time.scaled) # can be slow

rel.est <- with(fm.cls, Est/mean) # relative estimate
```

Here is what the first few results of `cls_star` look like:

```{r print cls_star output}
head(fm.cls)
```

We can see that site 1 has an average NDVI of 
`r round(fm.cls[1, "mean"], 2)` over the `r p` year time period. We can also 
see that, on average, NDVI decreased by `r round(abs(fm.cls[1, "Est"]), 2)` 
(`Est`) with a standard error of `r round(fm.cls[1, "SE"], 2)` (`SE`) over the 
entire time from 1982 to 2013. 

The t-test statistic and associated two-tailed p-value (`t` and `p`) indicate that
this change is significantly different from 0 (`p` $\leq .05 $. Finally, we can see that an 
increase in NDVI of 1 in the previous year is estimated to increase NDVI in 
the current year by `r round(fm.cls[1, "x_t0.EST"], 2)` at site 1 (`x_t0.EST`).

#### Check for outliers

Next we'll check for, and remove, outliers based on `rel.est`:

```{r plot AK3000 outl map}
# index of possible outliers
outl <- abs(scale(rel.est)) > -qnorm(p = 1/n/10) # criteria

## Make a plot to visualize where outliers are
# define color of points low NDVI = orange, high NDVI = green
base.col = ifelse(test = outl, yes = "black",
                  no = colorRampPalette(
                    c("darkgoldenrod3", "grey70", "chartreuse3")
                    )(nrow(fm.cls))[ordered(rel.est)]
)

# make legend labels from rel.est summary stats
labels <- round(c(max = max(rel.est), # boundaries of color changes
                  mid = median(rel.est), 
                  min = min(rel.est)), 2)
labels <- c(labels, outl = "outlier") # outliers

# build the plot
plot(y = ndvi_AK3000$lat, x = ndvi_AK3000$lng, pch = 15, cex = .5, 
     col = base.col,
     xlab = "longitude", ylab = "latitude")
legend(x = "bottomright", fill = c("chartreuse3", "grey70",
                                   "darkgoldenrod", "black"),
       legend = labels, title = "NDVI change")
```

We can see from the above figure that the outliers seem to occur at the edges of
the map - adjacent to missing data. This would be easier to see with the full
data set (`ndvi_AK`).

Let's see where they fall relative to the other data:

```{r outl hist}
## value of outliers relative to overall distribution
hist(fm.cls$mean[outl], freq = FALSE, breaks = 0:20, ylim = c(0,.5), col = "grey",
     lty = 3, xlab = "site average NDVI", main = "Histogram of NDVI")
hist(fm.cls$mean[!outl], freq = FALSE, breaks = 0:20, ylim = c(0,.5), col = NULL,
     add = TRUE)
legend("topright", legend = c("TRUE","FALSE"), fill = c("grey", "white"),
       title = "outlier")
```

This histogram doesn't seem to indicate that the outliers exhibit any pattern. 
However, this is not true for the full data set (**TBA: add a figure**)

For the purposes of this document, we will not remove the potential outliers 
but it may be a good idea depending on the dataset. 

### Using GLS regression

Now, we will use `remoteSTAR`'s GLS regression to test whether there is a 
significant trend in NDVI for this subset of Alaska:

The first step is to estimate the spatial correlation with `fit_spatialcor()`:

```{r fit_spatialcor}
# calculate distance matrix (km) to use later
D = geosphere::distm(location)/1000
meth = "exponential-power" # transformation method

# estimate spatial correlation
(r.est <- fit_spatialcor(X, time.scaled, location = location, 
                         fun = meth)) #calculates D internally
```

which gets used when fitting a covariance matrix by `fitV()`:

```{r fitV}
## Fit variance matrix
V <- fitV(D, spatialcor = r.est$spatialcor, fun = meth)
```

Then we can use `fitGLS_cpp()` to get our results. Here we also calculate some
values that `fitGLS_cpp` uses. 

**More info TBA**

```{r}
# get the inverse of the chol decomp of V (can be time consuming)
tInvCholV <- invert_chol(V, nugget = 0)

## get model matrix from formulae
# full model
form = "rel.est ~ land" 
mod.mat <- model.matrix(object = formula(form), data = ndvi_AK3000)
# null model
form0 = "rel.est ~ 1"
X0 <- model.matrix(object = formula(form0), data = ndvi_AK3000)

## estimate maximum likelihood nugget
tolr = .00001 # precision of nugget search
nugget.ml <- optimize_nugget(X = mod.mat, V = V, y = rel.est,
                                lower = 0, upper = 1, tol = tolr)

## fit GLS 
fm.gls <- fitGLS(X = mod.mat, V = V, y = rel.est, X0 = X0,
                     nugget = nugget.ml)
```

#### t-test

Now we'll use the results to formally test our hypotheses. 

First, we'll test $H_0:$ "land class *i* does not have a temporal trend in NDVI." 

```{r GLS t_pvalues}
# add in pvalues (this will eventually be done with classes internally)
fm.gls$pval.t <- sapply(fm.gls$tstat, function(x){
  2 * pt(abs(x), df = fm.gls$dft, lower.tail = F)
})
# feature names
names(fm.gls$pval.t) <- names(fm.gls$betahat) <- gsub(x = colnames(mod.mat),
                                                      pattern = "land", 
                                                      replacement = "")

# print p-values
cbind("pt" = fm.gls$pval.t) # fail to reject H0 for all land classes
```
We are unable to reject $H_0$ at $\alpha = .05$ for any land class in this 
data set. 

#### F-test

Now, let's test $H_0:$ "there is no overall effect of land class on NDVI trend"

```{r}
  # F-test
fm.gls$pval.F <- pf(fm.gls$Fstat, df1 = fm.gls$df.F[1], df2 = fm.gls$df.F[2], 
                    lower.tail = FALSE)

# ## print the coefficients
# round(fm.gls$betahat, 4)

# F-test ## H0: no change in NDVI over time for all of AK3000
round(cbind("F" = fm.gls$Fstat, "pval" = fm.gls$pval.F), 4) # reject H0
```

We are, however, able to reject this $H_0$ for this data set at $\alpha = .05$. 

#### Partitioned GLS

For extremely large data sets, computational constraints such as limited memory,
make it desirable to use a partitioned version of the GLS analysis. This version
of the method splits the data into `npart` random and non-overlapping sub-sets 
by site and then calculates cross-partition statistics on *nc* pairs of 
partitions (according to `mincross`) and summarizes the results with a 
correlated F-test.

The function that performs this analysis is the wrapper `fitGLS.partiion`. What
follows is a short example of how to perform this method with the `Alaska3000` 
data set but is not necessarily indicative of how users would want to use it 
in practice. See "More on the partitioned GLS method" for further clarification.

```{r partitioned GLS}
## run the parittioned analysis
results.cpp <- fitGLS.partition_rcpp(X = mod.mat, y = rel.est, X0 = X0, Dist = D,
                                 spatcor = r.est$spatialcor, Vfit.fun = meth,
                                 npart = 5, mincross = 4, nug.int = c(0, 1),
                                 nug.tol = tolr, workerB_cpp = TRUE)
```

Here are the F-test results for each partition:

```{r}
# F tests for each partition
t(sapply(results.cpp$part_results, function(x){c("F" = round(x$Fstat, 4), 
                                           "pval" = round(x$pval.F, 4))}))
```

and the summary F-test:

```{r}
# Cross-partition F test
pval.Fpart <- GLS.partition.pvalue(results.cpp, nboot = 2000)

round(cbind("Fmean" = results.cpp$Fmean, "pval" = pval.Fpart$pF.boot$pvalue), 4)
```


## More on the partitioned GLS method

Here, we will provide a description of how to use the individual functions for
the partitioned GLS method. 

First, we will use `sample_partitions()` to get a sample the pixels to use for each
subset. The result is a matrix with columns containing pixel IDs. Each column
corresponds to a different partition

```{r}
# segment dataset into 4 equal-sized partitions:
part.mat <- sample_partitions(nrow(mod.mat), npart = 4)
# part.mat <- sample_partitions(nrow(mod.mat), partsize = 1000)
```

Then, we will calculate the degrees of freedom with `calc_df()`, and perform 
GLS on each partition with `GLS_worker()`. The main arguments that `GLS_worker()`
needs are similar to those needed by `fitGLS()`: a response `y`, a model matrix
of predictors `X`, a null model matrix `X0` to compare against, and a variance 
matrix `V`. Each input is partition-specific. 

The difference between `GLS_worker()` and `fitGLS()` is that 
`fitGLS()` needs a nugget to be input (defaults to 0) while `GLS_worker()` finds
the maximum likelihood estimator of the nugget internally.

```{r}
# calculate degrees of freedom up front
dfs <- calc_df(nrow(part.mat), ncol(mod.mat), ncol(X0))
df1 <- dfs[1]
df2 <- dfs[2]

# make empty list to store the output
part.results <- list() 

# Use GLS worker function on each partition
for(part in seq_len(ncol(part.mat))){
  ## index setup
  subs = part.mat[, part] # current subset index
  X.sub = mod.mat[subs, ] # current subset of model matrix
  y.sub = rel.est[subs] # current subset of Y
  loc.sub = location[subs, ] # current coordinates
  X0.sub = X0[subs, ] # current subset of X0
  ## calculate variance matrix
  D.sub = geosphere::distm(loc.sub) # distance between subset points
  V.sub = fitV(D.sub, spatialcor = r.est$spatialcor, 
               fun = "exponential-power") # variance of subset points
  ## obtain GLS results
  part.results[[part]] <- GLS_worker(y.sub, X.sub, V.sub, X0.sub, save_xx = TRUE) 
}
```

The list `part.results` now has 4 elements - one for each partition - which are
GLS output lists (i.e. results for partition 1 are stored in 
`part.results[[1]]`).

Then, for each pair of GLS partition results, we can get the cross-partition
statistics with `crosspart_worker()`. This function takes as input (1) 
statistics returned by `GLS_worker()` (including the nugget estimate), (2) 
degrees of freedom, and (3) the cross-partition variance matrix `V.ij` which 
can be obtained from a matrix of distances between points from subsets `i` and 
`j`: 

```{r}
# make empty list for cross-partition results
cross.results = list()

# use crosspart_worker() to get cross-partition statistics
for(cross in seq_len(length(part.results) - 1)){ # each consecutive pair of partitions
  ## index setup
  i = cross # first partition
  j = cross + 1 # second partition
  subs.i = part.mat[, i] # index for part i
  subs.j = part.mat[, j] # index for part j
  ## GLS_worker() output
  Li = part.results[[i]] # list containing partition results i
  Lj = part.results[[j]] # list containing partition results j
  ## Obtain V.ij
  loc.i = location[subs.i, ] # coordinates of i
  loc.j = location[subs.j, ] # coordinates of j
  D.ij = geosphere::distm(loc.i, loc.j) # distance between i and j
  V.ij = fitV(D.ij, r.est$spatialcor) # variance matrix ij
  ## obtain cross-partition results
  cross.results[[cross]] = crosspart_worker(xxi = Li$xx, xxj = Lj$xx,
                                            xxi0 = Li$xx0, xxj0 = Lj$xx0,
                                            invChol_i = Li$invcholV,
                                            invChol_j = Lj$invcholV,
                                            nug_i = Li$nugget,
                                            nug_j = Lj$nugget,
                                            Vsub = V.ij,
                                            df1 = df1, df2 = df2)
}
```

`crosspart_worker()` is usually much faster than `GLS_worker()`.

Finally, we can calculate some average statistics for the entire analysis by
using `sapply()` to summarize across lists and then perform our hypothesis
test:

```{r}
## Average partition stats 
F.mean = mean(sapply(part.results, function(x){x$Fstat})) # average partition F
coef.means = rowMeans(sapply(part.results, function(x){x$betahat})) # coef est (land class)
coef0.mean = mean(sapply(part.results, function(x){x$betahat0})) # null coefs
part.SEs = sapply(part.results, function(x){x$SE}) # partition standard errors

## average cross-partition stats
rcoef.mean = rowMeans(sapply(cross.results, function(x){x$rcoef})) #cross-coef for conf ints
rSSR = mean(sapply(cross.results, function(x){x$rSSRij})) # regression sum of squares
rSSE = mean(sapply(cross.results, function(x){x$rSSEij})) # residual sum of squares


# calculate the correlated test statistics
## correlated chisquared test: are trends different among land class?
( p.chisqr = correlated.chisq(F.mean, rSSR, df1, npart = ncol(part.mat)) )
# correlated F test: are trends different among land class?
( p.corF = correlated.F.bootstrap(F.mean, rSSR, rSSE, df1, df2, 
                                  npart = ncol(part.mat), nboot = 2000) )
# t-test: are trends in land classes different from 0?
  (t.test = correlated.t(coefs = coef.means, part.SEs = part.SEs, rcoef = rcoef.mean, 
                   df2 = df2, npart = ncol(part.mat)) )
```

`fitGLS_partition()` does (**WILL**) all of this behind the scenes but it may be important
when analyzing extremely large data sets to know how the individual components
work together. The above steps could be easily extended for distributed or 
parallel computing which may be necessary if (A) the entire data set cannot be
processed in memory and/or (B) processing time is important and many individual 
computers or cores are available. 

For example, each subset partition (say, 100K pixels) could be run 
through `GLS_worker()` separately, output could be saved to individual files, 
memory could be cleaned or recycled at each step. Then, you could load the 
output files, 2 at a time, and run them through `crosspart_worker()`, saving the
cross-partition results to files and cleaning up memory after each pair. 

Average statistics (`F.mean`, `rSSR`, etc) could then be calculated by looping
through the files and keeping only the relevant statistics in memory (or writing
them to separate files). 

We will not discuss this type of parallelization or distributed computing in 
detail here but simply note that implementing this functionality could be 
trivially simple. That being said, minimal multi-core functionality is 
implemented internally through the C++ libraries `Eigen` and `openMP` (**NOTE** 
this is still not reliably present in the package). 

# Debugging remoteSTAR fitGLS_parititon functions

There are some discrepancies between my output and Tony's, so I'm going to try
to solve the problem. 

```{r}
## Check that the correlated t-test would lead to the same conclusions overall
  ## as the tests for individual partitions:

# coeficients
cfs = sapply(part.results, function(x)x$betahat)
rownames(cfs) = colnames(mod.mat)
colnames(cfs) = paste0("part",1:ncol(cfs))
# p values
pts = sapply(part.results, function(x){x$pval.t})
rownames(pts) = colnames(mod.mat)
colnames(pts) = paste0("part",1:ncol(pts))
# t scores
tscrs = sapply(part.results, function(x){x$tstat})
rownames(tscrs) = colnames(mod.mat)
colnames(tscrs) = paste0("part",1:ncol(tscrs))
# se
SEs = sapply(part.results, function(x){x$SE})
rownames(SEs) = colnames(mod.mat)
colnames(SEs) = paste0("part",1:ncol(SEs))

t.crit = qt(1 - (.05 / 2), df = df2 * ncol(part.mat))

library(ggplot2)

ggplot(data.frame(t.test, var = factor(colnames(mod.mat), levels = colnames(mod.mat))), 
       aes(x = var, y = coefs)) + 
  geom_hline(yintercept = 0, lty = 3, col = "black") + 
  geom_errorbar(aes(ymin = coefs - se*t.crit, 
                    ymax = coefs + se*t.crit, 
                    col = "correlated t-test", ), position = position_nudge(x = -.025),
                width = .1) +
  geom_errorbar(aes(ymin = coefs - apply(cfs, 1, sd)*t.crit, 
                    ymax = coefs + apply(cfs, 1, sd)*t.crit, 
                    col = "sd(coef)"),
                width = .1, position = position_nudge(x = .025)) + 
  geom_point() +
  labs(col = "95% CI", y = "coef", x = NULL, title = "H0: coef = 0")

```

```{r}
# need to test against tony's code
# source("")
dat.map = data.frame(rel.est, location, land = ndvi_AK3000$land)
library(geosphere)

form = "rel.est ~ land"

## Tony's code
Z <- GLS.partition.data(formula = "rel.est ~ land", 
                        formula0 = "rel.est ~ 1", 
                        data = dat.map,
                        spatial.autocor.FUN = "exponential-power",
                        spatialcor = r.est$spatialcor,
                        est.nugget = TRUE,
                        npart = 4,
                        partition = t(part.mat),
                        nugget.tol = 1e-05,
                        min.num.cross.part = 5,
                        verbose = FALSE,
                        rm.spatial.autocorrelation = FALSE
)

## nugget is not working properly
rbind(Z$nugget.part, 
      sapply(part.results, function(x)x$nugget))
## coefficient estimates are also wrong.
rbind(Z$coef, rowMeans(sapply(part.results, function(x)x$betahat)))

## not working with nugget = 0
# Z <- GLS.partition.data(formula = "rel.est ~ land", 
#                         formula0 = "rel.est ~ 1", 
#                         data = dat.map,
#                         spatial.autocor.FUN = "exponential-power",
#                         spatialcor = r.est$spatialcor,
#                         est.nugget = FALSE,
#                         fixed.nugget = 0,
#                         npart = 4,
#                         partition = t(part.mat),
#                         nugget.tol = 1e-05,
#                         min.num.cross.part = 5,
#                         verbose = FALSE,
#                         rm.spatial.autocorrelation = FALSE
# )

n1 = fitNugget(mod.mat, V, rel.est) # old from remoteSTAR
n2 = fitNugget_Rcpp(mod.mat, V, rel.est) # uses loglik function
n.tony = nugget.fit(form, data = dat.map, V = V) #tony's 
# n.parts = optimize_nugget(mod.mat, V, rel.est) # C++ version: not working at all

cbind(n1, n.tony)
```



```{r save alaska vignette data, eval = FALSE, echo = FALSE}
save(list = ls(), file = "alaska-vignette-data.RData")

load("alaska-vignette-data.RData", verbose = TRUE)
```

### `fitGLS_partition()`

**NOTE** This section still needs work

The previous example of `fitGLS_partiton()` was performed on the AK3000 dataset
which was loaded into memory. As stated earlier, this may not be feasible for 
large datasets. Luckily, `fitGLS_partition()` has functionality to address this
problem. 

but requires a 
RasterStack or CSV as input (to prevent loading the entire dataset into memory)




```{r}
## Test of using a function as an argument to a function

loc = location[sample(x = 1:nrow(location), size = 50), ]
loc2 = location[sample(x = 1:nrow(location), size = 50),]


D = distm(loc)

test.dist <- function(location, fun = "distm", ...){
  if(is.character(fun) && fun == "distm"){requireNamespace("geosphere")}
  fun <- match.fun(fun)
  D = fun(location, ...)
  return(D)
}


all(D == test.dist(loc))

D2 <- distm(loc, loc2)

D.test <- test.dist(loc, y = loc2, fun = function(x, y){
  distm(x, y)
})

all(D2 == D.test)

D3 <- dist(loc)
```


## Using Rasters

```{r, eval = FALSE}
## convert data to raster format
library(raster)
# make backbone
rast.backbone <- raster(xmn = min(location$lng), xmx = max(location$lng),
                        ymn = min(location$lat), ymx = max(location$lat),
                        ncol = length(unique(location$lng)),
                        nrow = length(unique(location$lat)),
                        # nrow = 1e4, ncol = 1e4 ## use 1e8 pixels
                        )

# initialize RasterStack
AK.stack <- stack()

# add layer for each column of X
for(i in 1:ncol(X)){
  # create raster from data and raster backbone
  AK.rast <- rasterize(as.matrix(location), rast.backbone, X[,i], fun = mean)
  AK.rast <- trim(AK.rast, internal = TRUE)
  names(AK.rast) <- colnames(X)[i]
  # add raster as layer to the stack
  AK.stack <- addLayer(AK.stack, AK.rast)
}

writeRaster(AK.stack, "../data/AK-RasterStack.grd", overwrite = TRUE)

AK.stack <- brick("../data/AK-RasterStack.grd")

canProcessInMemory(AK.stack)

land.class <- factor((mod.mat %*% (1:ncol(mod.mat))) + 1, 
         labels = c(colnames(mod.mat)))

land.raster <- rasterize(as.matrix(location), rast.backbone, as.numeric(land.class))
```

Since our raster files are mostly ocean, the grid is comprised of mostly empty cells:

```{r, eval = FALSE}
summary(AK.stack[[1]])
```


```{r, eval = FALSE}
# calculate the CLS coefficient for the raster file
CLS.rast = calc(AK.stack, 
                function(x){if(is.na(x[1])){NA
                }else{
                  fitCLS(x, t = 1:nlayers(AK.stack))$coef["time", "Estimate"]}
                })
## First, make a list of the data cells to include


# I'll simply do this by selecting the cells that are not NA in the first layer
dat.indx <- which(!is.na(values(AK.stack[[1]])))

part.mat <- sample_partitions(pixels = dat.indx, npart = 4)

# exctract pixels from first partition
rast.sub <- extract(AK.stack, part.mat[,1])


rast.sub[1:5, 1:5] #head(rast.sub)
y.sub = extract(CLS.rast, part.mat[,1])
land  = extract(land.raster, part.mat[, 1])
mm = model.matrix(~ 0 + factor(land))
colnames(mm) <- levels(land.class)

loc = xyFromCell(CLS.rast, cell = part.mat[, 1])

r.est = fit_spatialcor(X = rast.sub, t = 1:nlayers(AK.stack), location = loc)

D = geosphere::distm(loc)
V = fitV(D, spatialcor = r.est$spatialcor)
dfs <- calc_df(nrow(part.mat), p = ncol(X), p0 = ncol(X0))
## and then the rast.sub can be used just like a matrix in GLS
X0 = cbind(rep(1, length(y.sub)))

(out.GLS <- GLS_worker(y = y.sub, X = mm, V = V, X0 = X0))
```

```{r, eval = FALSE}
N = 2e8 #200Mill pixels

rasterfile = "../raster-test.grd"

tmp <- raster(ncol = sqrt(N), nrow = sqrt(N))

writeRaster(x = tmp, filename = rasterfile, overwrite = TRUE)

tmp <- raster(rasterfile)


tmp[50] <-  1

which(!is.na(values(tmp)))

extract(tmp, 50)
```

## Old Code / Test Code: 

```{r, echo = FALSE, eval = FALSE}
## shows that the correlated chisqr test is randombly distributed about 0
## when it is sufficiently low. 

F.means = seq(50, .1, length.out = 100)
p.vals <- sapply(F.means, function(x){
  correlated.chisq(x, rSSR, df1, ncol(part.mat))
})

plot(x = F.means[1:95], y = p.vals[1:95])
```

<!-- However, this dataset is very small and results can vary: lets look at a few -->
<!-- different subsets of `ndvi_AK`, each of 3000 sites. -->

```{r, eval = FALSE, echo = FALSE}
data("ndvi_AK")
ndvi_AK <- ndvi_AK[!ndvi_AK$rare.land, ]
ndvi_AK$land <- droplevels(ndvi_AK$land)

## 4 different subsets
subs <- matrix(sample.int(n = nrow(ndvi_AK), size = 3000 * 4, replace = FALSE), 
               nrow = 3000)

## Apply the test to 4 different subsets of 3000 AK sites
Ftest_sub <- sapply(1:ncol(subs), function(i){
  ss = sample.int(n = nrow(ndvi_AK), size = 3000, replace = FALSE)
  AKsub <- ndvi_AK[ss, ]
  loc_i <- AKsub[, c("lng", "lat")]
  Xsub <-  as.matrix(AKsub[, 7:38])
  sc_i <- fit_spatialcor(X = Xsub, t = time.scaled, 
                         location = loc_i, fun = meth)$spatialcor
  D_i <- geosphere::distm(loc_i)/1000
  fm.cls_i <- cls_star(Xsub, t = time.scaled)
  y_i <- with(fm.cls_i, Est/mean)
  fm_i <- fitGLS.partition_rcpp(X = Xsub, y = y_i, X0 = X0, Dist = D_i, 
                                spatcor = sc_i, Vfit.fun = meth, npart = 5, 
                                mincross = 4, workerB_cpp = TRUE)
  F_i <- fm_i$Fmean
  p_i <- GLS.partition.pvalue(fm_i, nboot = 2000)$pF.boot$pvalue
  
  return(c("F" = F_i, "pval" = p_i))
})

print(Ftest_sub)

rm("ndvi_AK")
```

<!-- I'm still working on the partitioned version -->

```{r GLS_partition, eval = FALSE, echo = FALSE}
n.p = 50; n.part = 3
partition <- matrix(sample(1:nrow(ndvi_AK3000), size = n.p * n.part), ncol = n.part)
x.1 = as.vector(partition[, 1])
x.2 = as.vector(partition[, 2])
x.12 = c(x.1, x.2)
## For now, I'm going to explore the distributed computing option

### partition 1
y1 <- rnorm(n.p)
X1 <- as.matrix(ndvi_AK3000[x.1, -c(1:6)])
loc1 <- ndvi_AK3000[x.1, c("lat","lng")]
V1 <- fitV(D[x.1, x.1],
            spatialcor = r.est$spatialcor, fun = "exponential-power")

### partition 2
y2 <- rnorm(n.p)
X2 <- as.matrix(ndvi_AK3000[x.2, -c(1:6)])
loc2 <- ndvi_AK3000[x.2, c("lat","lng")]
V2 <- fitV(D[x.2, x.2],
            spatialcor = r.est$spatialcor, fun = "exponential-power")

## cross partition
V12 <- fitV(D[x.12, x.12],
             spatialcor = r.est$spatialcor, fun = "exponential-power")
Vsub <- V12[1:n.p, (n.p+1):(2*n.p)]

Xnull <- matrix(1, nrow = nrow(X1))

df2 <- n.p - (ncol(X1) - 1)
df0 <- n.p - (ncol(Xnull) - 1)
df1 <- df0 - df2
```

```{r, eval = FALSE, echo = FALSE}
## results
out1 <- GLS_worker_cpp(y1, X1, V1, Xnull, save_xx = TRUE)
out2 <- GLS_worker_cpp(y2, X2, V2, Xnull, save_xx = TRUE)

## The cross-partition step is where it breaks down
## last worked when n.p was 400 but no larger
out.cross.cpp <- crosspart_worker_cpp(xxi = out1$xx, xxj = out2$xx,
                                  xxi0 = out1$xx0, xxj0 = out2$xx0,
                                  tUinv_i = out1$tInvCholV, 
                                  tUinv_j = out2$tInvCholV, 
                                  Vsub = Vsub, 
                                  df1 = df1, df2 = df2) # not working yet

out.cross <- crosspart_worker(xxi = out1$xx, xxj = out2$xx,
                                  xxi0 = out1$xx0, xxj0 = out2$xx0,
                                  tUinv_i = out1$tInvCholV, 
                                  tUinv_j = out2$tInvCholV, 
                                  Vsub = Vsub, 
                                  df1 = df1, df2 = df2) # not working yet

## test that the 2 versions give the same output
for(i in 1:length(out.cross)){
  print(names(out.cross)[i])
  print(all.equal(out.cross[[i]], out.cross.cpp[[i]]))
}


```

```{r, eval = FALSE, echo = FALSE}
size.Mb <- function(x){format(object.size(x), units = "Kb")}
sizes <- data.frame(obj = c("out1", "out2", "outcross"), 
                    MB = c(size.Mb(out1), size.Mb(out2), size.Mb(out.cross))
                    )
sizes

## could it be a memory issue? MB is the size of out.cross with different n.p values
curve = data.frame(n.p = c(100, 80, 60), MB = c(3.5, 2.2, 1.2))
fm <- lm(curve$MB ~ curve$n.p)
plot(curve$MB ~ curve$n.p); abline(fm)
pred <- function(np){(np * coef(fm)[2]) + coef(fm)[1]}
pred(c(150, 300, 600))
```

